PLAE enables quick and easy searching across one million ocular transcriptomes

### 2021-10-23 manuscript_00.Rmd
text

library(pool)
library(RSQLite)
meta_filter <- fst::read_fst("~/data/scEiaD_v2/2021_10_15_meta_filter.fst")
```

@@

b <-  meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  filter(Organ == 'Eye') %>% 
  group_by(Organ, Tissue,organism, study_accession) %>%
  summarise(Count = n()) %>%
  ungroup() %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  facet_wrap(~Organ, scales = 'free') +
  coord_flip()

# retina cell types
c <- meta_filter %>% filter(Organ == 'Eye') %>% 
  group_by(CellType_predict, study_accession, organism) %>%
  summarise(Count = n()) %>% filter(Count > 10) %>% ungroup() %>% 
  ggplot(aes(x=CellType_predict,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  coord_flip()

@@

## Update scEiaD resource
Our first version of scEiaD contained XX studies, 3 species, XX cells, and XX curated cell types. We updated the scEiaD resource for the PLAE web app in several important ways. First, we added two chicken (*Gallus gallus*) studies. Second, we added a human pan-body reference scRNA dataset. Third, we used an *in silico* method to remove background gene contamination. Fourth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained.

@@

>>>>

library(pool)
library(RSQLite)
meta_filter_v0 <- fst::read_fst("~/data/scEiaD/2021_03_17_meta_filter.fst")
# remove eric bharti 2D/3D RPE
meta_filter <- fst::read_fst("~/data/scEiaD_v2/2021_10_15_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D')

```

@@

# eye studies
b <-  meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  filter(Organ == 'Eye') %>% 
  group_by(Organ, Tissue,organism, study_accession) %>%
  summarise(Count = n()) %>%
  ungroup() %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  facet_wrap(~Organ, scales = 'free') +
  coord_flip()
# eye studies samples
c <-  meta_filter %>%
  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
                           TRUE ~ 'Body'),
         Tissue = gsub('\\.|_', ' ', Tissue),
         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
         Tissue = gsub(" and Aorta","", Tissue)) %>% 
  filter(Organ == 'Eye') %>% 
  select(Organ, Tissue,organism, study_accession) %>%
  unique() %>%
  group_by(Organ, Tissue, organism) %>% 
  summarise(Count = n()) %>%
  ungroup() %>% 
  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  facet_wrap(~Organ, scales = 'free') +
  coord_flip()

# retina cell types
d <- meta_filter %>% filter(Organ == 'Eye') %>% 
  group_by(CellType_predict, study_accession, organism) %>%
  summarise(Count = n()) %>% filter(Count > 10) %>% ungroup() %>% 
  ggplot(aes(x=CellType_predict,y=Count, fill = organism, group = organism)) +
  ylab('Number of Cells') +
  geom_bar(stat='identity', position = 'dodge') +
  cowplot::theme_cowplot() +
  coord_flip()

@@

@@

## Updated the scEiaD backend
Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types. We updated the scEiaD resource for the PLAE v0.80 web app in four important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset. Third, we added a human pan-body reference scRNA dataset. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX). Fourth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450.

After the dataset and quality control updates the scEiaD v02 dataset contains `r meta_filter_$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).

## PLAE contains 10 fold more cells and studies than the closest comparable web app

## Extensive optimization makes the PLAE web app highly responsive despite containing huge amounts of data
